---
title: "Plotting with R"
subtitle: "Descriptive analysis and basic statistics in biomedical studies using R and Markdown"
author: | 
  | Juan R Gonzalez
  | juanr.gonzalez@isglobal.org
institute: |
  | BRGE - Bioinformatics Research Group in Epidemiology
  | ISGlobal - Barcelona Institute for Global Health
  | http://brge.isglobal.org
date: |
  | IACS - Instituto Aragonés de Ciencias de la Salud
  | Zaragoza, February 26th
output:
  beamer_presentation:
    toc: false
    slide_level: 2
    includes:
      in_header: header.tex  
---


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,comment="", message=FALSE, warning=FALSE, cache=TRUE, fig.width = 4, fig.height = 4)
options(width=80)
```
 
# Plotting

## Plotting

- Simple plotting:
    * `plot`, `hist`, `pairs`, `boxplot`, ...
- Adding to existing plots:
    * `points`, `lines`, `abline`, `legend`, `title`, `mtext`, ...
- Interacting with graphics:
    * `locator`, `identify`
- Three dimensional data:
    * `contour`, `image`, `persp`, ...
- To see the many possibilities that R offers:
```{r eval=FALSE}
demo(graphics)
```

--- 

Basic plotting function is `plot()`. Possible arguments to
`plot()` include:

* x, y (y may be omitted)
* xlim = c(lo, hi), ylim = c(lo, hi)
* `xlab = "x"`, `ylab = "y"` -- labels for x- and y-axes respectively
* `type = "c"` -- type of plot ("p", "l", "b", "h", "S", ...)
* "lty = n" -- line type (if lines used)
* "lwd = n" -- line width
* "pch = v" -- plotting character(s)
* "col = v" -- colour to be used for everything.

---

```{r}
library(MASS)
head(Cars93)
```

---

```{r}
plot(Cars93$Weight, Cars93$EngineSize)
```

---


```{r}
# Colour points and choose plotting symbols according to 
# a levels of a factor
plot(Cars93$Weight, Cars93$EngineSize, 
     col=as.numeric(Cars93$Type),
     pch=as.numeric(Cars93$Type))
```

--- 


```{r}
# Add x and y axes labels and a title.
plot(Cars93$Weight, Cars93$EngineSize, ylab="Engine Size",
xlab="Weight", main="My plot")
```

---

```{r}
plot(Cars93$Weight, Cars93$EngineSize, ylab="Engine Size",
xlab="Weight", main="My plot")

# Add lines to the plot.
lines(x=c(min(Cars93$Weight), max(Cars93$Weight)),
      y=c(min(Cars93$EngineSize),
      max(Cars93$EngineSize)), lwd=4, lty=3, col="green")
abline(h=3, lty=2)
abline(v=1999, lty=4)
```

--- 

```{r}
plot(Cars93$Weight, Cars93$EngineSize, ylab="Engine Size",
xlab="Weight", main="My plot")

# Add points to the plot
points(x=min(Cars93$Weight), y=min(Cars93$EngineSize), pch=16, col="red")
```


---


```{r plotLegend, fig.show='hide'}
plot(Cars93$Weight, Cars93$EngineSize, ylab="Engine Size",
xlab="Weight", main="My plot")

# Add text to the plot.
text(x=2000, y=5, "some text")

# Add text under main title.
mtext(side=3, "sub-title", line=0.45)

# Add a legend
legend("bottomright", legend=c("Data Points"), pch="o")
```

--- 

\includegraphics[width=3in]{Day3_plotting_files/figure-beamer/plotLegend-1}

--- 

```{r}
plot(Cars93$Weight, Cars93$EngineSize, ylab="Engine Size",
xlab="Weight", main="My plot")

# Add regression line
mod <- lm(EngineSize ~ Weight, data=Cars93)
abline(mod)
````

--- 

```{r regOrgin, fig.show='hide'}
levels(Cars93$Origin)
plot(Cars93$Weight, Cars93$EngineSize, 
     pch = (1:2)[Cars93$Origin],
     col = (2:3)[Cars93$Origin], 
     xlab="Weight", ylab="Engine Size")
legend("topleft", legend=levels(Cars93$Origin), 
       pch=1:2, col=2:3)

fm1 <- lm(EngineSize ~ Weight, Cars93, subset = Origin == "USA")
abline(coef(fm1), lty=4, col="blue")

fm2 <- lm(EngineSize ~ Weight, Cars93, subset = Origin == "non-USA")
abline(coef(fm2), lty=4, col="black")
```
--- 

\includegraphics[width=3in]{Day3_plotting_files/figure-beamer/regOrgin-1}


# Multiple figures

---

```{r multiple, fig.show='hide'}
# Will create 4 plots on the same page.
# Two in each row and two in each column.
par(mfrow=c(2,2))       

plot(Cars93$Weight, Cars93$EngineSize, 
     xlab="Weight", ylab="Engine Size")

plot(sqrt(Cars93$Weight), sqrt(Cars93$EngineSize),
     xlab=expression(sqrt(Weight)), 
     ylab=expression(sqrt("Engine Size")))

plot(log(Cars93$Weight), log(Cars93$EngineSize),
     xlab=expression(log(Weight)), 
     ylab=expression(log("Engine Size")))

plot(Cars93$Weight^0.1, Cars93$EngineSize^0.1,
     xlab=expression(Weight^0.1), 
     ylab=expression("Engine Size"^0.1) )

par(mfrow=c(1,1))     # Resets to create a single plot per page.
```

**[Advanced:]** The `expression` command plots mathematical symbols on the x and y axes. For more information `help(expression)

---

\includegraphics[width=3in]{Day3_plotting_files/figure-beamer/multiple-1}

--- 

Improve the figure by

```{r multiple2, fig.show='hide'}
par("mar")
par(mar=c(5,4,1,2))
par(mfrow=c(2,2))       
plot(Cars93$Weight, Cars93$EngineSize, 
     xlab="Weight", ylab="Engine Size")
plot(sqrt(Cars93$Weight), sqrt(Cars93$EngineSize),
     xlab=expression(sqrt(Weight)), 
     ylab=expression(sqrt("Engine Size")))
plot(log(Cars93$Weight), log(Cars93$EngineSize),
     xlab=expression(log(Weight)), 
     ylab=expression(log("Engine Size")))
plot(Cars93$Weight^0.1, Cars93$EngineSize^0.1,
     xlab=expression(Weight^0.1), 
     ylab=expression("Engine Size"^0.1) )
```

---

\includegraphics[width=3in]{Day3_plotting_files/figure-beamer/multiple2-1}

---

## Histograms
Histograms can be created using the `hist` command. Let us create a histogram of the car weights from the Cars93 data set:

```{r}
hist(Cars93$Weight, xlab="Weight", 
     main="Histogram of Weight", col="red")
```

---

R automatically chooses the number and width of the bars. Can change this by specifying the number of break points. 

```{r}
hist(Cars93$Weight, breaks=8, xlab="Weight", 
     main="Histogram of Weight", col="blue")
```

---

Normal curve can be added to the histogram by:

```{r histnorm}
y <- Cars93$Weight # put here your variable of interest
hist(y, freq=FALSE)
lines(density(y), col="red")
x <- seq(min(y), max(y), length=100)
lines(x, dnorm(x, mean(y), sd(y)), col="blue")
```
 

## Histograms for multiple groups

```{r histgroups, fig.show='hide'}
USA.weight <- Cars93$Weight[Cars93$Origin == "USA"]
nonUSA.weight <- Cars93$Weight[Cars93$Origin == "non-USA"]

par(mfrow=c(2,1))
par(mar=c(5,4,1,2))
hist(USA.weight, breaks=10, xlim=c(1500,4500), col="grey")
hist(nonUSA.weight, breaks=10, xlim=c(1500,4500))
par(mfrow=c(1,1))
```
---
  
\includegraphics[width=3in]{Day3_plotting_files/figure-beamer/histgroups-1}  


## Boxplots

```{r}
boxplot(Cars93$Weight)
```

--- 

```{r}
boxplot(Cars93$Weight ~ Cars93$Origin, 
        col="blue", ylab="Weight")
```

---

## Normal probability (Q-Q) plots

```{r}
qqnorm(Cars93$Weight)
qqline(Cars93$Weight)
```

## Plots for multivariate data
If your data are stored in a data frame with several columns,
the `pairs` command produces pairwise plots of the data
in each column, i.e. the data in column 1 vs the data in column 2,
column 1 vs column 3, and so on.

```{r}
pairs(Cars93[,1:7])
```

--- 

This is limited to 2 grouping variables. More flexible
is `xyplot` in the `lattice` library.

```{r}
coplot(Cars93$MPG.city~Cars93$Price|Cars93$Type)
```

## Lattice graphs

```{r}
library(lattice)
xyplot(Cars93$MPG.city~Cars93$EngineSize|Cars93$Type)
```

---

```{r}
xyplot(Cars93$MPG.city~Cars93$EngineSize|Cars93$Type*Cars93$Origin)
```

---

```{r}
xyplot(Cars93$MPG.city~Cars93$EngineSize|Cars93$Type,
       panel=panel.superpose, groups=Cars93$Origin)
```

---

```{r}
# Adds a key
pars <- trellis.par.get("superpose.symbol")
xyplot(Cars93$MPG.city~Cars93$EngineSize|Cars93$Type,
       panel=panel.superpose, groups=Cars93$Origin, 
       key = list(columns = 2, text = list(levels(Cars93$Origin)),
                               points = Rows(pars,1:2)))
```

## Other lattice plots

* splom( ~ data.frame) # Scatterplot matrix
* bwplot(factor ~ numeric, ...) # Boxplot
* qqmath(factor ~ numeric, ...) # Q-Q plot
* dotplot(factor ~ numeric, ...) # 1-D display
* stripplot(factor ~ numeric, ...)
* barchar(character ~ numeric, ...)
* histogram( ~ numeric, ...)
* densityplot( ~ numeric, ...) # Smoothed version of histogram


## 2-D and 3-D plots

```{r plot2d, fig.show='hide'}
data(volcano)
x <- 10*(1:nrow(volcano))
y <- 10*(1:ncol(volcano))

# Creates a 2-D image of x and y co-ordinates.
image(x, y, volcano, col = terrain.colors(100),
      axes = FALSE)

# Adds contour lines to the current plot.
contour(x, y, volcano, levels = seq(90, 200, by=5),
        add = TRUE, col = "peru")

# Adds x and y axes to the plot.
axis(1, at = seq(100, 800, by = 100))
axis(2, at = seq(100, 600, by = 100))

# Draws a box around the plot.
box()

# Adds a title.
title(main = "Maunga Whau Volcano", font.main = 4)
```

---

\includegraphics[width=3in]{Day3_plotting_files/figure-beamer/plot2d-1}

---




## Session info

```{r}
sessionInfo()
```

